---
title: "ICFES Saber 11"
subtitle: "Munging data"
author: "Miguel Martinez"
date: today
format: 
  html:
    toc: true
    theme: cosmo
editor: visual
execute:
  warning: false
---

This document presents the process about importing, tranforming and filtering the SABER 11 test from 2010-2022 in Colombia. You can download the data [here](https://www.datos.gov.co/Educaci-n/Resultados-nicos-Saber-11/kgxf-xxbe/about_data).

The first step is to load the following library:

```{r}

library(tidyverse)
```

## Load data

Now we can load the data:

```{r}

data = read.csv("https://www.datos.gov.co/resource/kgxf-xxbe.csv")

glimpse(data)
```

## Transformation

In the following code, we transform the `PERIODO` variable into:

1.  *Year*

2.  *Period*

For example: the value 20131 in the `PERIODO` variable, means that the exam was made in the first period of the year 2013.

```{r}

data = data |>
  separate_wider_position(periodo,
                          widths = c(ANO = 4,
                                     PER = 1))

glimpse(data)
```

## Examen Saber 11 - 2019

This is the Saber 11 test filtered for the year 2019.

```{r}

data2019 = data |>
  filter(ANO == "2019")

glimpse(data2019)

```

## Examen Saber 11 - 2022

This is the Saber 11 test filtered for the year 2022.

```{r}

data2022 = data |>
  filter(ANO == "2022")

glimpse(data2022)

```

## Examen Saber 11 - 2019 y 2022

This is the Saber 11 test filtered for the year 2019 and 2022.

```{r}

data1922 = data |>
  filter(ANO %in% c("2019", "2022"))

glimpse(data1922)

```

## Exporting the data

To export the data frames, you just need to uncomment and run the following code chunks:

```{r}

# write.csv(data2019, file = 'saber11_2019.csv', row.names = FALSE, quote = FALSE, sep = ",")
```

```{r}

# write.csv(data2022, file = 'saber11_2022.csv', row.names = FALSE, quote = FALSE, sep = ",")
```

```{r}

# write.csv(data1922, file = 'saber11_2019_2022.csv', row.names = FALSE, quote = FALSE, sep = ",")
```
